---
title: "Differential Loop Analysis"
author: "JP Flores"
format: html
editor: visual
execute: 
  eval: false
---

#### Load necessary packages

```{r}
#| warning: false
library(DESeq2)
library(dplyr)
library(glue)
library(stringr)
library(purrr)
library(pheatmap)
library(apeglm)
```

#### Load [extracted loops](https://jpflores.quarto.pub/generating-loop-counts/).


```{r}
loopCounts <- readRDS("~/Phanstiel Lab Dropbox/JP Flores/projects/YAPP/YAPP/data/processed/hic/YAPP_hic_loopCounts.rds")
```

Adding column `loop_name` to metadata.

```{r}
## add a loop_count column
loopCounts$loop_name <- glue("loop_{1:length(loopCounts)}")
```

### Assembling the count matrix

From the metadata cols of `loopCounts`, pull out the `inter_30.hic` files and create a matrix.

```{r}
m <- mcols(loopCounts)[, grep("*inter.*", colnames(mcols(loopCounts)))] %>% 
  as.matrix()

tail(m)
```

### Construct colData/metadata table

Using the information from our count matrix, we split the column names of the matrix at the underscore. This creates a list of 6 where each split string turns into a column name. We then bind the rows of each list to create a matrix. Finally, this matrix is converted into a `data.frame`.

The column names of our count matrix ultimately turns into the row names of our metadata table (colData). The column names are specified (Project, Cell Type, Treatment, and Replicate) and explicitly called.

```{r}
## String split the colnames 
colData <- as.data.frame(do.call(rbind, strsplit(colnames(m), "_")), stringsAsFactors = T)
rownames(colData) <- colnames(m)
colnames(colData) <- c("Project", "Cell_Type", "Treatment", "Replicate")
colData <- colData[, c(1:4)]

## Correct replicate number
colData <- colData |> 
  mutate(Replicate = str_replace(Replicate, "4", "1")) |> 
  mutate(Replicate = str_replace(Replicate, "5", "2")) |> 
  mutate(Replicate = str_replace(Replicate, "6", "3"))

print(colData)
```

In order to confirm that the column names of our count matrix is equal to the row names of our metadata table, we can run an `all()` function.

```{r}
## Make sure sample names match
all(colnames(m) == rownames(colData))
```

## Create a DESeq object

Because we created matrices, we use the `DESeqDataSetFromMatrix()` function. We also need to specify a design formula. The design formula specifies the column(s) in the metadata table and how they should be used in the analysis. A design formula tells the statistical software the known sources of variation to control for, as well as, the factor of interest to test for during differential expression testing. The design formula should have all of the factors in your metadata that account for major sources of variation in your data. The last factor entered in the formula should be the condition of interest.

```{r}
#| warning: false
dds <- DESeqDataSetFromMatrix(countData = m,
                       colData = colData,
                       design = ~ Replicate + Treatment)
```

Rather than using `DESeq()`'s default normalization method (Median of ratios method), we can turn this off by setting the size factors of our `dds` object manually. Here, we set the size factors of each column to 1.

:::{.callout-important}
THIS IS THE DIFFERENCE MAKER!
:::

```{r}
sizeFactors(dds) <- rep(1, ncol(dds))
```

After turning normalization off, we can then use the `DESeq()` function. Here, we will try using both the Wald test and LRT test.

```{r}
#| warning: false
## Hypothesis testing with Wald with `betaPrior = F`
dds <- DESeq(dds)
```


```{r}
#| results: hide
#| warning: false
## results from Wald test
res <- results(dds)
resultsNames(dds)
res <- lfcShrink(dds, coef="Treatment_sorbitol_vs_control", type= "apeglm")
```

```{r}
summary(res)
```

The next step in the workflow is to fit a curve to the dispersion estimates for each gene. The idea behind fitting a curve to the data is that different genes will have different scales of biological variability, but, over all genes, there will be a distribution of reasonable estimates of dispersion.

This curve is displayed as a red line in the figure below, which plots the estimate for the expected dispersion value for genes of a given expression strength. Each black dot is a gene with an associated mean expression level and maximum likelihood estimation (MLE) of the dispersion (Step 1).

```{r}
## Plot dispersion estimates for Wald Test results
plotDispEsts(dds)
```

### MA plots

```{r}
plotMA(res, ylim=c(-4,4), main = "Differential Loops Analysis - Wald",
       ylab = "LFC",
       xlab = "mean of norm. counts")
```


### Other QC metrics


Principal Component Analysis
```{r}
## plot PCA
plotPCA(vst(dds), intgroup = "Treatment") + ggplot2::theme(aspect.ratio = 1)

## PCA results
plotPCA(vst(dds), intgroup = "Treatment", returnData = TRUE)
```


Hierarchical Clustering
```{r}
## transform counts for hierarchical clustering
rld <- rlog(dds, blind=TRUE)

## Extract the rlog matrix from the object
rld_mat <- assay(rld)

## Compute pairwise correlation values
rld_cor <- cor(rld_mat)

## Plot heatmap
pheatmap(rld_cor)
```


### Concatenate loopCounts and DESeqResults
```{r}
mcols(loopCounts) <- cbind(mcols(loopCounts), res)
diff_loopCounts <- loopCounts
```


### Save as `.rds` and share session info
```{r}
saveRDS(diff_loopCounts, file = "data/processed/hic/YAPP_hic_diff_loopCounts.rds")

sessionInfo()
```








